---
title: "Step 1: Define Gene Groups"
author: "Daniel Aldas-Bulos"
date: "2025-07-21"
output: html_document
---

## Purpose

Define the gene groups used throughout the Figure 5 analysis:

- **g3**: Genes significantly upregulated (logFC ≥ 0.5, padj ≤ 0.05) by *both*
  gRNA2 and gRNA7 at 4 hpf upon cth1 KD.
- **c1–c4**: Four non-overlapping random control sets (n = 500 each) drawn from
  genes that are *not* significantly upregulated by either gRNA.
- **cth1**: The cth1 gene itself, highlighted separately.

Filtering criteria applied before group assignment:

1. 3'UTR length between 100 and 800 nt.
2. Expression ≥ 5 CPM in cas13d-alone controls.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source("00_setup.R")
```

## Load DGE results

```{r load-data}
original_KD <- fread(file.path(data_dir, "original_KD_DGE_results.csv"))

# Remove g13 columns (deprecated guide)
original_KD <- original_KD |>
  dplyr::select(Gene_ID, !contains("g13"))

DGE_OE <- fread(file.path(data_dir, "DGE_new_OE.csv"))

DGE_OE_4h <- DGE_OE |>
  dplyr::select(Gene_ID, Name, contains("4h") & !contains("orf_oe"))
```

## Apply filters

```{r apply-filters}
# 3'UTR length filter
len_3utr <- fread(file.path(data_dir, "utr3_protein_coding_longest_tx_Ariel.txt"))
len_3utr_filter <- len_3utr |> filter(Len <= 800 & Len >= 100)

# Expression filter (CPM >= 5 in cas13d-alone)
expression_filter <- original_KD |> filter(s_cas13d_2h1.CPM >= 5)

# Merge KD and OE at 4h, then apply both filters
DGE_results_4h_all <- merge(original_KD, DGE_OE_4h, by = "Gene_ID")

DGE_results_4h_filtered <- DGE_results_4h_all |>
  filter(Gene_ID %in% len_3utr_filter$Gene_ID &
         Gene_ID %in% expression_filter$Gene_ID)

DGE_OE_4h_filtered <- DGE_OE_4h |>
  filter(Gene_ID %in% DGE_results_4h_filtered$Gene_ID)
```

## Define g3 group (upregulated by both gRNAs)

```{r define-g3}
new_g3 <- DGE_results_4h_filtered |>
  filter(
    s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.logFC >= 0.5 &
    s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.padj  <= 0.05 &
    s_cas13d_pos_syn_g7_4h_VS_s_cas13d_4h.logFC >= 0.5 &
    s_cas13d_pos_syn_g7_4h_VS_s_cas13d_4h.padj  <= 0.05
  ) |>
  mutate(group = "g3")

cat("g3 group size:", nrow(new_g3), "genes\n")
```

## Define control groups (c1–c4)

```{r define-controls}
set.seed(42)

DGE_results_control <- DGE_results_4h_filtered |>
  filter(
    s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.logFC < 0.5 &
    s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.padj  > 0.05 &
    s_cas13d_pos_syn_g7_4h_VS_s_cas13d_4h.logFC < 0.5 &
    s_cas13d_pos_syn_g7_4h_VS_s_cas13d_4h.padj  > 0.05
  )

shuffled_indices <- sample(nrow(DGE_results_control))

c1 <- DGE_results_control[shuffled_indices[1:500], ]       |> mutate(group = "c1")
c2 <- DGE_results_control[shuffled_indices[501:1000], ]    |> mutate(group = "c2")
c3 <- DGE_results_control[shuffled_indices[1001:1500], ]   |> mutate(group = "c3")
c4 <- DGE_results_control[shuffled_indices[1501:2000], ]   |> mutate(group = "c4")

control_data <- bind_rows(c1, c2, c3, c4)

cat("Control pool size:", nrow(DGE_results_control), "genes\n")
cat("Each control group: 500 genes\n")
```

## Highlight data (for scatter plots)

```{r highlight-data}
cth1_data <- DGE_results_4h_filtered |>
  filter(Name %in% "cth1") |>
  mutate(group = "cth1")

highlight_data <- bind_rows(new_g3, cth1_data)
```

## Build groups lookup table

```{r groups-table}
groups <- bind_rows(
  new_g3 |> mutate(group = "g3"),
  c1     |> mutate(group = "c1"),
  c2     |> mutate(group = "c2"),
  c3     |> mutate(group = "c3"),
  c4     |> mutate(group = "c4")
) |>
  dplyr::select(Gene_ID, group)

groups <- left_join(groups, len_3utr |> dplyr::select(Gene_ID, Gene_Name, Transcript_ID), by = "Gene_ID")

cat("Total genes in groups table:", nrow(groups), "\n")
```

## Save intermediate objects

```{r save}
# Save key objects for use by downstream scripts
save(
  original_KD, DGE_OE_4h, DGE_results_4h_filtered, DGE_OE_4h_filtered,
  len_3utr, len_3utr_filter, expression_filter,
  new_g3, c1, c2, c3, c4, control_data, highlight_data, cth1_data,
  groups,
  file = file.path(results_dir, "01_groups.RData")
)
```
