---
title: "Step 2: KD Scatter Plots and Overlap Diagrams"
author: "Daniel Aldas-Bulos"
date: "2025-07-21"
output: html_document
---

## Purpose

Visualize the concordance between gRNA2 and gRNA7 knockdown effects:

- Scatter plots of log2 FC (gRNA2 vs gRNA7) with gene groups highlighted.
- Scatter plots for alpha-amanitin–treated conditions.
- Euler/Venn diagrams showing overlap of significantly upregulated and
  downregulated genes across gRNAs and timepoints.
- Pure zygotic gene overlays.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source("00_setup.R")
load(file.path(results_dir, "01_groups.RData"))
```

## Significance classification

```{r significance-classification}
original_KD <- original_KD |>
  mutate(significant_in = case_when(
    (s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.padj <= 0.05 &
     s_cas13d_pos_syn_g7_4h_VS_s_cas13d_4h.padj <= 0.05) ~ "Both",
    (s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.padj <= 0.05 &
     s_cas13d_pos_syn_g7_4h_VS_s_cas13d_4h.padj >  0.05) ~ "g2",
    (s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.padj >  0.05 &
     s_cas13d_pos_syn_g7_4h_VS_s_cas13d_4h.padj <= 0.05) ~ "g7",
    .default = "none"
  ))

original_KD$significant_in <- factor(
  original_KD$significant_in,
  levels = c("none", "g2", "g7", "Both")
)
```

## gRNA2 vs gRNA7 scatter (significance)

```{r scatter-significance}
scatter_significant <- original_KD |>
  arrange(significant_in) |>
  ggplot(aes(
    x     = s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.logFC,
    y     = s_cas13d_pos_syn_g7_4h_VS_s_cas13d_4h.logFC,
    color = significant_in,
    alpha = significant_in
  )) +
  geom_point() +
  scale_color_manual(values = c("lightgray", "palegreen3", "lightsalmon", "dodgerblue")) +
  scale_alpha_manual(values = c(0.3, 0.5, 0.5, 0.8)) +
  theme_minimal()

scatter_significant
ggsave(file.path(figures_dir, "scatter_significant.pdf"), scatter_significant, width = 7, height = 7)
```

## gRNA2 vs gRNA7 scatter (gene groups)

```{r scatter-groups}
kd_scatter_groups <- DGE_results_4h_filtered |>
  ggplot(aes(
    x = s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.logFC,
    y = s_cas13d_pos_syn_g7_4h_VS_s_cas13d_4h.logFC
  )) +
  geom_point(fill = alpha("lightgray", 0.5), color = "lightgray",
             shape = 21, stroke = 0.15, size = 0.8) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(data = control_data,
             aes(fill = group, color = group),
             shape = 21, stroke = 0.15, size = 0.8) +
  ggpubr::stat_cor(color = "black", size = 3) +
  geom_point(data = highlight_data,
             aes(fill = group, color = group),
             shape = 21, stroke = 0.3, size = 0.8) +
  geom_text_repel(
    data = highlight_data |> filter(group == "cth1"),
    aes(label = group),
    segment.colour = "black", segment.size = 0.3,
    segment.alpha = 0.7, min.segment.length = 0.1
  ) +
  guides(color = "none",
         fill = guide_legend(override.aes = list(shape = 21, stroke = 0.3, size = 4))) +
  scale_fill_manual(values = alpha(colors_guide, 0.5)) +
  scale_color_manual(values = colors_guide) +
  coord_cartesian(xlim = c(-2, 3), ylim = c(-2, 3)) +
  theme_pubr() +
  labs(x = "RNA Log2 FC cth1 KD with gRNA2 vs cas13d alone",
       y = "RNA Log2 FC cth1 KD with gRNA7 vs cas13d alone",
       fill = "Group")

kd_scatter_groups
ggsave(file.path(figures_dir, "scatter_KD_groups.pdf"), kd_scatter_groups, width = 7, height = 7)
```

## Alpha-amanitin scatter

```{r scatter-alpha}
control_data_alpha <- merge(
  control_data |> dplyr::select(Gene_ID, group),
  DGE_OE_4h_filtered,
  by = "Gene_ID"
)

highlight_data_alpha <- merge(
  highlight_data |> dplyr::select(Gene_ID, group),
  DGE_OE_4h_filtered,
  by = "Gene_ID"
)
highlight_data_alpha <- rbind(
  highlight_data_alpha,
  DGE_OE_4h |> filter(Name == "cth1") |> mutate(group = "cth1")
)

scatter_alpha <- DGE_OE_4h_filtered |>
  ggplot(aes(
    x = s_cas13d_alfa_g2_4h_VS_s_cas13d_alfa_alone_4h.logFC,
    y = s_cas13d_alfa_g7_4h_VS_s_cas13d_alfa_alone_4h.logFC
  )) +
  geom_point(fill = alpha("lightgray", 0.5), color = "lightgray",
             shape = 21, stroke = 0.15, size = 0.8) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(data = control_data_alpha,
             aes(fill = group, color = group),
             shape = 21, stroke = 0.15, size = 0.8) +
  ggpubr::stat_cor(color = "black", size = 3) +
  geom_point(data = highlight_data_alpha,
             aes(fill = group, color = group),
             shape = 21, stroke = 0.3, size = 0.8) +
  geom_text_repel(
    data = highlight_data_alpha |> filter(group == "cth1"),
    aes(label = group),
    segment.colour = "black", segment.size = 0.3,
    segment.alpha = 0.7, min.segment.length = 0.1
  ) +
  guides(color = "none",
         fill = guide_legend(override.aes = list(shape = 21, stroke = 0.3, size = 4))) +
  scale_fill_manual(values = alpha(colors_guide, 0.5)) +
  scale_color_manual(values = colors_guide) +
  coord_cartesian(xlim = c(-2, 3), ylim = c(-2, 3)) +
  theme_pubr() +
  labs(x = "RNA Log2 FC cth1 KD with gRNA2 vs cas13d alone (alpha-amanitin)",
       y = "RNA Log2 FC cth1 KD with gRNA7 vs cas13d alone (alpha-amanitin)",
       fill = "Group")

scatter_alpha
ggsave(file.path(figures_dir, "scatter_KD_groups_alpha.pdf"), scatter_alpha, width = 7, height = 7)
```

## Euler diagrams (upregulated / downregulated gene overlap)

```{r euler-upregulated}
# Helper: extract upregulated gene sets
extract_upreg <- function(df, lfc_col, padj_col, label) {
  df |>
    filter(.data[[lfc_col]] > 0, .data[[padj_col]] <= 0.05) |>
    transmute(Gene_ID, grna = label, value = TRUE)
}

upregulated <- bind_rows(
  extract_upreg(original_KD, "s_cas13d_pos_syn_g2_2h_VS_s_cas13d_2h.logFC",
                "s_cas13d_pos_syn_g2_2h_VS_s_cas13d_2h.padj", "sg2 2hpf"),
  extract_upreg(original_KD, "s_cas13d_pos_syn_g7_2h_VS_s_cas13d_2h.logFC",
                "s_cas13d_pos_syn_g7_2h_VS_s_cas13d_2h.padj", "sg7 2hpf"),
  extract_upreg(original_KD, "s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.logFC",
                "s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.padj", "sg2 4hpf"),
  extract_upreg(original_KD, "s_cas13d_pos_syn_g7_4h_VS_s_cas13d_4h.logFC",
                "s_cas13d_pos_syn_g7_4h_VS_s_cas13d_4h.padj", "sg7 4hpf")
) |>
  distinct(Gene_ID, grna, .keep_all = TRUE) |>
  pivot_wider(names_from = grna, values_from = value,
              values_fill = list(value = FALSE), values_fn = list(value = any))

fit_up <- euler(upregulated |> dplyr::select(-Gene_ID))

p_up <- plot(fit_up, shape = "ellipse", ggplot = TRUE, quantities = TRUE,
             labels = FALSE, legend = TRUE, main = "Upregulated genes",
             fills = list(fill = c("sg2 2hpf" = "darkolivegreen1",
                                   "sg7 2hpf" = "coral1",
                                   "sg2 4hpf" = "darkolivegreen4",
                                   "sg7 4hpf" = "coral4"), alpha = 0.6))
p_up
ggsave(file.path(figures_dir, "venn_upregulated.png"), p_up, limitsize = FALSE, width = 7, height = 7)
```

```{r euler-downregulated}
extract_downreg <- function(df, lfc_col, padj_col, label) {
  df |>
    filter(.data[[lfc_col]] < 0, .data[[padj_col]] <= 0.05) |>
    transmute(Gene_ID, grna = label, value = TRUE)
}

downregulated <- bind_rows(
  extract_downreg(original_KD, "s_cas13d_pos_syn_g2_2h_VS_s_cas13d_2h.logFC",
                  "s_cas13d_pos_syn_g2_2h_VS_s_cas13d_2h.padj", "sg2 2hpf"),
  extract_downreg(original_KD, "s_cas13d_pos_syn_g7_2h_VS_s_cas13d_2h.logFC",
                  "s_cas13d_pos_syn_g7_2h_VS_s_cas13d_2h.padj", "sg7 2hpf"),
  extract_downreg(original_KD, "s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.logFC",
                  "s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.padj", "sg2 4hpf"),
  extract_downreg(original_KD, "s_cas13d_pos_syn_g7_4h_VS_s_cas13d_4h.logFC",
                  "s_cas13d_pos_syn_g7_4h_VS_s_cas13d_4h.padj", "sg7 4hpf")
) |>
  distinct(Gene_ID, grna, .keep_all = TRUE) |>
  pivot_wider(names_from = grna, values_from = value,
              values_fill = list(value = FALSE), values_fn = list(value = any))

fit_down <- euler(downregulated |> dplyr::select(-Gene_ID))

p_down <- plot(fit_down, shape = "ellipse", ggplot = TRUE, quantities = TRUE,
               labels = FALSE, legend = TRUE, main = "Downregulated genes",
               fills = list(fill = c("sg2 2hpf" = "darkolivegreen1",
                                     "sg7 2hpf" = "coral1",
                                     "sg2 4hpf" = "darkolivegreen4",
                                     "sg7 4hpf" = "coral4"), alpha = 0.6))
p_down
ggsave(file.path(figures_dir, "venn_downregulated.png"), p_down, limitsize = FALSE, width = 7, height = 7)
```

## Pure zygotic gene overlay

```{r zygotic-scatter}
pure_zygotic <- fread(file.path(data_dir, "Zygotic.csv"))

zygotic_data <- DGE_OE_4h_filtered |>
  filter(Gene_ID %in% pure_zygotic$gene_id) |>
  mutate(group = "Pure zygotic genes")

zygotic_scatter <- DGE_OE_4h_filtered |>
  ggplot(aes(
    x = s_cas13d_alfa_g2_4h_VS_s_cas13d_alfa_alone_4h.logFC,
    y = s_cas13d_alfa_g7_4h_VS_s_cas13d_alfa_alone_4h.logFC
  )) +
  geom_point(fill = alpha("lightgray", 0.5), color = "lightgray",
             shape = 21, stroke = 0.15, size = 0.8) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(data = zygotic_data,
             aes(fill = group, color = group),
             shape = 21, stroke = 0.15, size = 0.8) +
  guides(color = "none",
         fill = guide_legend(override.aes = list(shape = 21, stroke = 0.3, size = 4))) +
  scale_fill_manual(values = alpha("gold", 0.5)) +
  scale_color_manual(values = "gold") +
  theme_pubr() +
  labs(x = "RNA Log2 FC cth1 KD with gRNA2 vs cas13d alone (alpha-amanitin)",
       y = "RNA Log2 FC cth1 KD with gRNA7 vs cas13d alone (alpha-amanitin)",
       fill = "Group")

zygotic_scatter
ggsave(file.path(figures_dir, "zygotic_scatter.pdf"), zygotic_scatter, width = 7, height = 7)
```

## Save intermediate objects

```{r save}
save(
  scatter_alpha, control_data_alpha, highlight_data_alpha,
  file = file.path(results_dir, "02_scatter.RData")
)
```
