---
title: "Step 4: RNA Decay Analysis"
author: "Daniel Aldas-Bulos"
date: "2025-07-21"
output: html_document
---

## Purpose

Compute RNA decay rates as log2 fold change of TPM at later timepoints
relative to 2 hpf, for three library types: RiboZero, Poly(A)-selected,
and alpha-amanitin–treated Poly(A). Compare decay between g3 and control
groups using boxplots, sina plots, violin plots, ECDFs, and Wilcoxon tests.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source("00_setup.R")
load(file.path(results_dir, "01_groups.RData"))
load(file.path(results_dir, "03_timecourse.RData"))
```

## Helper: compute decay log2 FC

```{r helper}
#' Compute log2 FC of TPM at later timepoints vs. a reference timepoint
#' @param df         Data.frame with TPM columns.
#' @param prefix     Column prefix pattern (e.g., "Treatment_wt-").
#' @param suffix     Column suffix (e.g., "_hrs-RNAseq_ribo").
#' @param ref_tp     Reference timepoint string (e.g., "2").
#' @param later_tps  Character vector of later timepoints.
#' @param label      Label prefix for output columns (e.g., "Ribo").
compute_decay <- function(df, prefix, suffix, ref_tp, later_tps, label) {
  pseudocount <- 0.05
  ref_col <- paste0(prefix, ref_tp, suffix)
  for (tp in later_tps) {
    tp_col  <- paste0(prefix, tp, suffix)
    out_col <- paste0(label, "_", tp, "hpf/", ref_tp, "hpf")
    df[[out_col]] <- log2((df[[tp_col]] + pseudocount) / (df[[ref_col]] + pseudocount))
  }
  df
}
```

## i. RiboZero decay

```{r decay-ribo}
decay_ribo <- santi |>
  filter(`Treatment_wt-0_hrs-RNAseq_ribo` > 5) |>
  dplyr::select(Gene_ID, ends_with("_ribo")) |>
  compute_decay("Treatment_wt-", "_hrs-RNAseq_ribo", "2", 3:8, "Ribo") |>
  dplyr::select(-starts_with("Treatment")) |>
  pivot_longer(cols = starts_with("Ribo"),
               names_to = "timepoint", names_pattern = "Ribo_(.*)",
               values_to = "log2_fc")

decay_ribo_groups <- left_join(
  groups_timecourse |> dplyr::select(Gene_ID, group),
  decay_ribo, by = "Gene_ID"
)
```

```{r ribo-sina, fig.width=8, fig.height=6}
decay_ribo_sina <- decay_ribo_groups |>
  na.omit() |>
  ggplot(aes(x = timepoint, y = log2_fc, color = group, fill = group)) +
  geom_sina(shape = 21, size = 1, stroke = 0.15) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5,
               position = position_dodge(width = 0.9),
               aes(group = group), color = "black") +
  coord_cartesian(ylim = c(-3, 1)) +
  theme_pubr() +
  scale_fill_manual(values = alpha(colors_guide, 0.5)) +
  scale_color_manual(values = colors_guide) +
  labs(x = "Timepoint", y = "Log2 FC (late / 2 hpf)",
       title = "RNA decay — RiboZero", color = "Group")

decay_ribo_sina
ggsave(file.path(figures_dir, "decay_ribo_sina.pdf"), decay_ribo_sina)
```

```{r ribo-ecdf, fig.width=10, fig.height=6}
decay_ribo_ecdf <- decay_ribo_groups |>
  na.omit() |>
  ggplot(aes(x = log2_fc, color = group)) +
  stat_ecdf(geom = "step") +
  coord_cartesian(xlim = c(-3, 1)) +
  theme_pubr() +
  scale_color_manual(values = colors_guide) +
  labs(x = "Log2 FC (late / 2 hpf)",
       title = "RNA decay ECDF — RiboZero", color = "Group") +
  facet_wrap(~timepoint)

decay_ribo_ecdf
ggsave(file.path(figures_dir, "decay_ribo_ecdf.pdf"), decay_ribo_ecdf)
```

```{r ribo-pvalues}
ribo_pvals <- map_dfr(c("3hpf/2hpf", "4hpf/2hpf", "5hpf/2hpf", "6hpf/2hpf"), function(tp) {
  wilcox_vs_baseline(
    data.frame(decay_ribo_groups) |> na.omit() |> filter(timepoint == tp),
    baseline = "g3", value_col = "log2_fc", class_col = "group"
  ) |> mutate(timepoint = tp, baseline = "g3")
})

fwrite(ribo_pvals, file.path(results_dir, "pval_decay_ribo_groups.csv"))
```

## ii. Poly(A) decay

```{r decay-polya}
decay_polya <- santi |>
  filter(`Treatment_wt-0_hrs-RNAseq_ribo` > 5) |>
  dplyr::select(Gene_ID, !contains("aamanitin") & ends_with("_polya")) |>
  compute_decay("Treatment_wt-", "_hrs-RNAseq_polya", "2", 3:8, "Poly") |>
  dplyr::select(-starts_with("Treatment")) |>
  pivot_longer(cols = starts_with("Poly"),
               names_to = "timepoint", names_pattern = "Poly_(.*)",
               values_to = "log2_fc")

decay_poly_groups <- left_join(
  groups_timecourse |> dplyr::select(Gene_ID, group),
  decay_polya, by = "Gene_ID"
)
```

```{r polya-sina, fig.width=8, fig.height=6}
decay_poly_sina <- decay_poly_groups |>
  na.omit() |>
  ggplot(aes(x = timepoint, y = log2_fc, color = group, fill = group)) +
  geom_sina(shape = 21, size = 1, stroke = 0.15) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5,
               position = position_dodge(width = 0.9),
               aes(group = group), color = "black") +
  coord_cartesian(ylim = c(-3, 3)) +
  theme_pubr() +
  scale_fill_manual(values = alpha(colors_guide, 0.5)) +
  scale_color_manual(values = colors_guide) +
  labs(x = "Timepoint", y = "Log2 FC (late / 2 hpf)",
       title = "RNA decay — Poly(A)", color = "Group")

decay_poly_sina
ggsave(file.path(figures_dir, "decay_polya_sina.pdf"), decay_poly_sina)
```

```{r polya-pvalues}
polya_pvals <- map_dfr(c("3hpf/2hpf", "4hpf/2hpf", "5hpf/2hpf", "6hpf/2hpf"), function(tp) {
  wilcox_vs_baseline(
    data.frame(decay_poly_groups) |> na.omit() |> filter(timepoint == tp),
    baseline = "g3", value_col = "log2_fc", class_col = "group"
  ) |> mutate(timepoint = tp)
})

fwrite(polya_pvals, file.path(results_dir, "pval_decay_polya_groups.csv"))
```

## iii. Alpha-amanitin decay

```{r decay-alpha}
pseudocount <- 0.05

decay_alpha <- santi |>
  filter(`Treatment_wt-0_hrs-RNAseq_ribo` > 5) |>
  dplyr::select(Gene_ID, contains("aamanitin")) |>
  mutate(
    `Alpha_2.5hpf/2hpf` = log2((`Treatment_aamanitin-2.5_hrs-RNAseq_polya` + pseudocount) /
                                (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_3hpf/2hpf`   = log2((`Treatment_aamanitin-3_hrs-RNAseq_polya` + pseudocount) /
                                (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_3.5hpf/2hpf` = log2((`Treatment_aamanitin-3.5_hrs-RNAseq_polya` + pseudocount) /
                                (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_4hpf/2hpf`   = log2((`Treatment_aamanitin-4_hrs-RNAseq_polya` + pseudocount) /
                                (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_4.5hpf/2hpf` = log2((`Treatment_aamanitin-4.5_hrs-RNAseq_polya` + pseudocount) /
                                (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_5hpf/2hpf`   = log2((`Treatment_aamanitin-5_hrs-RNAseq_polya` + pseudocount) /
                                (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_5.5hpf/2hpf` = log2((`Treatment_aamanitin-5.5_hrs-RNAseq_polya` + pseudocount) /
                                (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_6hpf/2hpf`   = log2((`Treatment_aamanitin-6_hrs-RNAseq_polya` + pseudocount) /
                                (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_6.5hpf/2hpf` = log2((`Treatment_aamanitin-6.5_hrs-RNAseq_polya` + pseudocount) /
                                (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_7hpf/2hpf`   = log2((`Treatment_aamanitin-7_hrs-RNAseq_polya` + pseudocount) /
                                (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_7.5hpf/2hpf` = log2((`Treatment_aamanitin-7.5_hrs-RNAseq_polya` + pseudocount) /
                                (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount))
  ) |>
  dplyr::select(-starts_with("Treatment")) |>
  pivot_longer(cols = starts_with("Alpha"),
               names_to = "timepoint", names_pattern = "Alpha_(.*)",
               values_to = "log2_fc")

decay_alpha_groups <- left_join(
  groups_timecourse |> dplyr::select(Gene_ID, group),
  decay_alpha, by = "Gene_ID"
) |> na.omit()

decay_alpha_groups$timepoint <- factor(
  decay_alpha_groups$timepoint,
  levels = unique(decay_alpha_groups$timepoint)
)
```

```{r alpha-sina, fig.width=10, fig.height=6}
decay_alpha_sina <- decay_alpha_groups |>
  ggplot(aes(x = factor(timepoint), y = log2_fc, color = group, fill = group)) +
  geom_sina(shape = 21, size = 1, stroke = 0.15) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5,
               position = position_dodge(width = 0.9),
               aes(group = group), color = "black") +
  coord_cartesian(ylim = c(-2.5, 2)) +
  theme_pubr() +
  scale_fill_manual(values = alpha(colors_guide, 0.5)) +
  scale_color_manual(values = colors_guide) +
  labs(x = "Timepoint", y = "Log2 FC (late / 2 hpf)",
       title = "RNA decay — alpha-amanitin treated", color = "Group") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

decay_alpha_sina
ggsave(file.path(figures_dir, "decay_alpha_sina.pdf"), decay_alpha_sina)
```

```{r alpha-pvalues}
alpha_pvals <- map_dfr(c("3hpf/2hpf", "4hpf/2hpf", "5hpf/2hpf", "6hpf/2hpf"), function(tp) {
  wilcox_vs_baseline(
    data.frame(decay_alpha_groups) |> na.omit() |> filter(timepoint == tp),
    baseline = "g3", value_col = "log2_fc", class_col = "group"
  ) |> mutate(timepoint = tp)
})

fwrite(alpha_pvals, file.path(results_dir, "pval_decay_alpha_groups.csv"))
```

## Poly(A) status p-values

```{r polystatus-pvalues}
polya_status_data <- groups_timecourse |>
  dplyr::select(!contains("aamanitin")) |>
  mutate(
    PolyvsRibo_0h = log2((`Treatment_wt-0_hrs-RNAseq_polya` + 0.05) /
                         (`Treatment_wt-0_hrs-RNAseq_ribo` + 0.05)),
    PolyvsRibo_1h = log2((`Treatment_wt-1_hrs-RNAseq_polya` + 0.05) /
                         (`Treatment_wt-1_hrs-RNAseq_ribo` + 0.05)),
    PolyvsRibo_2h = log2((`Treatment_wt-2_hrs-RNAseq_polya` + 0.05) /
                         (`Treatment_wt-2_hrs-RNAseq_ribo` + 0.05)),
    PolyvsRibo_3h = log2((`Treatment_wt-3_hrs-RNAseq_polya` + 0.05) /
                         (`Treatment_wt-3_hrs-RNAseq_ribo` + 0.05)),
    PolyvsRibo_4h = log2((`Treatment_wt-4_hrs-RNAseq_polya` + 0.05) /
                         (`Treatment_wt-4_hrs-RNAseq_ribo` + 0.05))
  ) |>
  dplyr::select(-starts_with("Treatment")) |>
  pivot_longer(cols = starts_with("PolyvsRibo"),
               names_to = "timepoint", names_pattern = "PolyvsRibo_(.*)",
               values_to = "log2_fc")

polystatus_pvals <- map_dfr(c("0h", "1h", "2h", "3h", "4h"), function(tp) {
  wilcox_vs_baseline(
    data.frame(polya_status_data) |> na.omit() |> filter(timepoint == tp),
    baseline = "g3", value_col = "log2_fc", class_col = "group"
  ) |> mutate(timepoint = paste0(tp, "pf"), baseline = "g3")
})

fwrite(polystatus_pvals, file.path(results_dir, "polystatus_pval_all.csv"))
```

## Save intermediate objects

```{r save}
save(
  decay_ribo_groups, decay_poly_groups, decay_alpha_groups,
  polya_status_data,
  file = file.path(results_dir, "04_decay.RData")
)
```
