---
title: "Step 5: Characterize Gene Groups (CDS and UTR Lengths)"
author: "Daniel Aldas-Bulos"
date: "2025-07-21"
output: html_document
---

## Purpose

Query Ensembl (v106) via biomaRt to retrieve CDS, 3'UTR, and 5'UTR sequences
for all genes in the groups table, then compare length distributions between g3
and control groups using sina plots and Wilcoxon tests.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source("00_setup.R")
load(file.path(results_dir, "01_groups.RData"))
```

## Query biomaRt (Ensembl v106)

Using the April 2022 archive host (Ensembl release 106) and batched queries
to avoid HTTP 500 errors on heavy sequence attributes.

```{r biomart-query}
# I have to downgrade httr2 due to problems using bioMart
# remotes::install_version("httr2", version = "1.0.7")

ensembl <- biomaRt::useEnsembl(biomart = "genes", dataset = "drerio_gene_ensembl", version = 106)

cds <- batch_getBM(
  attributes = c("ensembl_transcript_id", "transcript_length", "coding"),
  filters    = "ensembl_transcript_id",
  values     = groups$Transcript_ID,
  mart       = ensembl
)

utr3 <- batch_getBM(
  attributes = c("ensembl_transcript_id", "transcript_length", "3utr"),
  filters    = "ensembl_transcript_id",
  values     = groups$Transcript_ID,
  mart       = ensembl
)

utr5 <- batch_getBM(
  attributes = c("ensembl_transcript_id", "transcript_length", "5utr"),
  filters    = "ensembl_transcript_id",
  values     = groups$Transcript_ID,
  mart       = ensembl
)

# Compute lengths
utr3 <- utr3 |> mutate(len_3utr = nchar(`3utr`))
utr5 <- utr5 |> mutate(len_5utr = nchar(`5utr`))
cds  <- cds  |> mutate(cds_len   = nchar(coding))
```

## Merge length data with groups

```{r merge-lengths}
group_lengths <- merge(utr3 |> dplyr::select(-`3utr`),
                       utr5 |> dplyr::select(-`5utr`)) |>
  merge(cds |> dplyr::select(-coding)) |>
  merge(groups, by.y = "Transcript_ID", by.x = "ensembl_transcript_id") |>
  mutate(
    tx_len     = len_5utr + cds_len + len_3utr,
    len_5_test = transcript_length - cds_len - len_3utr
  )
```

## CDS length distribution

```{r cds-sina, fig.width=7, fig.height=5}
cds_sina <- ggplot(group_lengths, aes(x = group, y = cds_len, fill = group, color = group)) +
  geom_sina(shape = 21, size = 1, stroke = 0.15) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5,
               position = position_dodge(width = 0.9),
               aes(group = group), color = "black") +
  coord_cartesian(ylim = c(0, 4900)) +
  theme_pubr() +
  scale_fill_manual(values = alpha(colors_guide, 0.5)) +
  scale_color_manual(values = colors_guide) +
  labs(x = "Group", color = "Group", y = "CDS length (nt)",
       title = "CDS length distribution")

cds_sina
ggsave(file.path(figures_dir, "cds_sinaplot.pdf"), cds_sina, width = 7, height = 7)
```

## 3'UTR length distribution

```{r utr3-sina, fig.width=7, fig.height=5}
sina_3utr <- ggplot(group_lengths, aes(x = group, y = len_3utr, fill = group, color = group)) +
  geom_sina(shape = 21, size = 1, stroke = 0.15) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5,
               position = position_dodge(width = 0.9),
               aes(group = group), color = "black") +
  theme_pubr() +
  scale_fill_manual(values = alpha(colors_guide, 0.5)) +
  scale_color_manual(values = colors_guide) +
  labs(x = "Group", color = "Group", y = "3'UTR length (nt)",
       title = "3'UTR length distribution")

sina_3utr
ggsave(file.path(figures_dir, "3utr_sinaplot.pdf"), sina_3utr, width = 7, height = 7)

len_3utr_pval <- wilcox_vs_baseline(
  group_lengths, class_col = "group", baseline = "g3", value_col = "len_3utr"
) |> mutate(baseline = "g3") |> dplyr::relocate(baseline, .before = everything())

fwrite(len_3utr_pval, file.path(results_dir, "len_3utr_pval.csv"))
```

## 5'UTR length distribution

```{r utr5-sina, fig.width=7, fig.height=5}
sina_5utr <- ggplot(group_lengths, aes(x = group, y = len_5_test, fill = group, color = group)) +
  geom_sina(shape = 21, size = 1, stroke = 0.15) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5,
               position = position_dodge(width = 0.9),
               aes(group = group), color = "black") +
  coord_cartesian(ylim = c(0, 500)) +
  theme_pubr() +
  scale_fill_manual(values = alpha(colors_guide, 0.5)) +
  scale_color_manual(values = colors_guide) +
  labs(x = "Group", color = "Group", y = "5'UTR length (nt)",
       title = "5'UTR length distribution")

sina_5utr
ggsave(file.path(figures_dir, "5utr_sinaplot.pdf"), sina_5utr, width = 7, height = 7)
```

## Save intermediate objects

```{r save}
# UTR sequences are needed downstream for motif analysis
utr_seqs <- merge(utr3, groups, by.x = "ensembl_transcript_id", by.y = "Transcript_ID") |>
  dplyr::rename(utr3 = `3utr`)

save(
  group_lengths, utr_seqs, utr3, utr5, cds, ensembl,
  file = file.path(results_dir, "05_characterize.RData")
)
```
