---
title: "Step 6: K-mer Enrichment and Meta-Analysis"
author: "Daniel Aldas-Bulos"
date: "2025-07-21"
output: html_document
---

## Purpose

Identify enriched RNA motifs (5–10 nt k-mers) in the 3'UTRs of g3 genes
compared to each of the four control groups (c1–c4). Combine p-values across
control comparisons using Fisher's method. Visualize enriched k-mers with
volcano plots, sequence logos, and positional bias analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source("00_setup.R")
 load(file.path(results_dir, "01_groups.RData"))
load(file.path(results_dir, "05_characterize.RData"))
```

## Prepare 3'UTR sequences by group

```{r prepare-seqs}
g3_utr <- utr_seqs |> filter(group == "g3")
c1_utr <- utr_seqs |> filter(group == "c1")
c2_utr <- utr_seqs |> filter(group == "c2")
c3_utr <- utr_seqs |> filter(group == "c3")
c4_utr <- utr_seqs |> filter(group == "c4")

g3_seqs <- DNAStringSet(g3_utr$utr3)
c1_seqs <- DNAStringSet(c1_utr$utr3)
c2_seqs <- DNAStringSet(c2_utr$utr3)
c3_seqs <- DNAStringSet(c3_utr$utr3)
c4_seqs <- DNAStringSet(c4_utr$utr3)

cat("g3 UTRs:", length(g3_seqs), "\n")
cat("Control UTRs: c1 =", length(c1_seqs), ", c2 =", length(c2_seqs),
    ", c3 =", length(c3_seqs), ", c4 =", length(c4_seqs), "\n")
```

## Export FASTA files (for external k-mer counting)

```{r export-fasta}
fasta_dir <- file.path("..", "data", "fasta")
dir.create(fasta_dir, showWarnings = FALSE, recursive = TRUE)

create_fasta(g3_utr, file.path(fasta_dir, "g3_3utr.fa"), as_rna = TRUE)
create_fasta(c1_utr, file.path(fasta_dir, "c1_3utr.fa"), as_rna = TRUE)
create_fasta(c2_utr, file.path(fasta_dir, "c2_3utr.fa"), as_rna = TRUE)
create_fasta(c3_utr, file.path(fasta_dir, "c3_3utr.fa"), as_rna = TRUE)
create_fasta(c4_utr, file.path(fasta_dir, "c4_3utr.fa"), as_rna = TRUE)
```

## Load k-mer enrichment results (from external pipeline)

The k-mer enrichment was computed on the HPC cluster comparing g3 vs each
control group. Results are read and combined here.

```{r load-kmers}
kmer_files <- list(
  c1 = file.path(results_dir, "kmers", "g3_all_vs_c1.csv"),
  c2 = file.path(results_dir, "kmers", "g3_all_vs_c2.csv"),
  c3 = file.path(results_dir, "kmers", "g3_all_vs_c3.csv"),
  c4 = file.path(results_dir, "kmers", "g3_all_vs_c4.csv")
)

kmers_raw <- imap_dfr(kmer_files, ~ fread(.x) |> data.frame() |> mutate(iteration = .y))

# Split by k-mer size
kmers_by_k <- map(5:10, ~ kmers_raw |> filter(k == .x))
names(kmers_by_k) <- paste0("kmers_", 5:10)
```
```{r}
fread("../results/kmers/g3_all_vs_c4.csv")
```


## Meta-analysis (Fisher's method)

```{r meta-analysis}
kmers_meta <- map(kmers_by_k, meta_analysis)
names(kmers_meta) <- paste0("kmers_", 5:10, "_p")

# Also run on all k-mers combined
kmers_all_p <- meta_analysis(kmers_raw)
```

## Volcano plots of enriched k-mers

```{r volcano-function}
enriched_kmers_plot <- function(df_plot, k = 5, xlim = c(-2, 2)) {

  # Subsample for plotting efficiency if very large
  if (nrow(df_plot) > 20000) {
    df_plot <- rbind(
      df_plot |> sample_n(15000),
      df_plot |> filter(fisher_padj < 1)
    )
  }

  enriched_data <- df_plot |> filter(fisher_padj <= 0.01, log2fc > 0)

  volcano <- ggplot(df_plot, aes(x = log2fc, y = -log10(fisher_padj))) +
    geom_hline(yintercept = -log10(0.01), color = "lightgray", linetype = 2) +
    geom_point(shape = 21, color = "lightgray", alpha = 0.5,
               stroke = 0.15, size = 0.8, fill = "lightgray") +
    geom_point(data = enriched_data,
               shape = 21, alpha = 0.5, stroke = 0.15, size = 1.2,
               fill = alpha("dodgerblue", 0.5), color = "dodgerblue") +
    geom_text_repel(data = enriched_data, aes(label = kmer), size = 2) +
    coord_cartesian(xlim = xlim) +
    theme_pubr() +
    labs(title = paste0("Enriched ", k, "-mers in g3 vs control groups"),
         x = "Log2 FC normalized k-mer counts (g3 / mean controls)",
         y = "-log10 p-value (Fisher's method)")

  ggsave(file.path(figures_dir, paste0("volcano_", k, "mers_g3.pdf")),
         volcano, width = 7, height = 7)

  return(volcano)
}
```

```{r plot-volcanoes, fig.width=7, fig.height=7}
enriched_kmers_plot(kmers_meta$kmers_5_p,  k = 5)
enriched_kmers_plot(kmers_meta$kmers_6_p,  k = 6)
enriched_kmers_plot(kmers_meta$kmers_7_p,  k = 7)
enriched_kmers_plot(kmers_meta$kmers_8_p,  k = 8, xlim = c(-2, 2.5))
enriched_kmers_plot(kmers_meta$kmers_9_p,  k = 9)
enriched_kmers_plot(kmers_meta$kmers_10_p, k = 10)
```

## Positional bias of enriched motifs

```{r positional-bias, fig.width=10, fig.height=5}
candidate_motifs <- c("AUAAAUA", "AAUAAAU")

# Find motif positions in each group
position_data <- map_dfr(
  list(g3 = g3_utr, c1 = c1_utr, c2 = c2_utr, c3 = c3_utr, c4 = c4_utr),
  function(utr_df) {
    find_motif_positions(utr_df, candidate_motifs) |>
      merge(utr_df |> dplyr::select(ensembl_transcript_id, len_3utr),
            by.x = "Transcript_ID", by.y = "ensembl_transcript_id")
  },
  .id = "group"
) |>
  mutate(
    relative_position = end / len_3utr,
    distance          = len_3utr - end
  )

# Distance to poly(A) tail
distance_plot <- position_data |>
  ggplot(aes(x = -distance, color = group)) +
  geom_density() +
  scale_color_manual(values = colors_guide) +
  theme_minimal() +
  labs(x = "Distance to poly(A) tail (nt)",
       color = "Group",
       title = "Motif distance to poly(A) tail") +
  facet_grid(~motif)

distance_plot
ggsave(file.path(figures_dir, "distance_plot.pdf"), distance_plot, width = 7, height = 7)

# Relative position along 3'UTR
relative_plot <- position_data |>
  ggplot(aes(x = relative_position, color = group)) +
  geom_density() +
  scale_color_manual(values = colors_guide) +
  theme_minimal() +
  labs(x = "Relative position along 3'UTR",
       color = "Group",
       title = "Motif relative position in 3'UTR") +
  facet_grid(~motif)

relative_plot
ggsave(file.path(figures_dir, "relative_position_plot.pdf"), relative_plot, width = 7, height = 7)
```

## Save intermediate objects

```{r save}
save(
  g3_utr, c1_utr, c2_utr, c3_utr, c4_utr,
  kmers_meta, kmers_all_p, position_data,
  file = file.path(results_dir, "06_motif.RData")
)
```
