---
title: "Step 7: Endogenous Motif Classification and KD Effects"
author: "Daniel Aldas-Bulos"
date: "2025-07-21"
output: html_document
---

## Purpose

Classify all filtered genes (not just g3) by the presence of enriched motifs
in their 3'UTRs, then compare KD effects and decay rates across motif classes.
Classes are defined hierarchically: 3A 8-mer > 2A 8-mer > more than one site > one site > rest.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source("00_setup.R")
load(file.path(results_dir, "01_groups.RData"))
load(file.path(results_dir, "03_timecourse.RData"))
load(file.path(results_dir, "05_characterize.RData"))
load(file.path(results_dir, "06_motif.RData"))
```

## Retrieve all 3'UTR sequences for filtered genes

```{r get-all-utrs}
transcripts_all <- len_3utr |> filter(Gene_ID %in% DGE_results_4h_filtered$Gene_ID)

# Reuse the ensembl connection from step 05 (loaded via RData),
# or reconnect if running this step standalone:
if (!exists("ensembl")) {
  ensembl <- useEnsembl(
    biomart = "genes",
    dataset = "drerio_gene_ensembl",
    host    = "https://apr2022.archive.ensembl.org"
  )
}

utr3_all <- batch_getBM(
  attributes = c("ensembl_transcript_id", "ensembl_gene_id", "transcript_length", "3utr"),
  filters    = "ensembl_transcript_id",
  values     = transcripts_all$Transcript_ID,
  mart       = ensembl
) |>
  dplyr::rename(utr3 = `3utr`, Gene_ID = ensembl_gene_id)

```

## Find motif occurrences (8-mers and 7-mers)

```{r find-motifs}
genes_8mer <- find_motif_positions(utr3_all, c("AAUAAAUA", "AAUAAUAA")) |>
  merge(utr3_all, by.x = "Transcript_ID", by.y = "ensembl_transcript_id") |>
  mutate(Len = nchar(utr3))

# Count motif types per transcript
classes_8mer <- genes_8mer |>
  group_by(Transcript_ID) |>
  summarise(
    n_AAA = sum(motif == "AAUAAAUA"),
    n_AA  = sum(motif == "AAUAAUAA"),
    .groups = "drop"
  )

AAA_transcripts <- classes_8mer |> filter(n_AAA > 0) |> distinct(Transcript_ID)
AA_transcripts  <- classes_8mer |> filter(n_AA > 0)  |> distinct(Transcript_ID)
just_1    <- classes_8mer |> mutate(total = n_AAA + n_AA) |> filter(total == 1) |> distinct(Transcript_ID)
more_than1 <- classes_8mer |> mutate(total = n_AAA + n_AA) |> filter(total > 1) |> distinct(Transcript_ID)
```

## Build endogenous motif groups

```{r build-groups}
endogenous_groups <- rbind(
  AAA_transcripts |> mutate(group = "3A 8mer"),
  AA_transcripts  |> mutate(group = "2A 8mer"),
  just_1          |> mutate(group = "one site"),
  more_than1      |> mutate(group = "more than one site"),
  utr3_all |> transmute(Transcript_ID = ensembl_transcript_id, group = "Rest of genes")
)

endogenous_groups_data <- merge(
  endogenous_groups,
  utr3_all |> dplyr::transmute(Gene_ID, Transcript_ID = ensembl_transcript_id),
  by = "Transcript_ID"
)
```

## Stacked bar chart of motif class distribution

```{r stacked-bar, fig.width=6, fig.height=5}
endogenous_count <- endogenous_groups |>
  dplyr::count(group, sort = TRUE) |>
  mutate(
    group = factor(group, levels = c("3A 8mer", "2A 8mer",
                                     "more than one site", "one site",
                                     "Rest of genes")),
    type = ifelse(group == "Rest of genes", "Without motif", "With motif")
  )

endogenous_stack <- ggplot(endogenous_count, aes(fill = group, y = n, x = type)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5), hjust = -0.3) +
  scale_fill_manual(values = highlight_colors) +
  theme_pubr() +
  labs(x = "", fill = "Group")

endogenous_stack
ggsave(file.path(figures_dir, "stack_endogenous_with_motif.pdf"), endogenous_stack, width = 7, height = 7)
```

## ECDF of KD effects by motif class

```{r kd-ecdf, fig.width=8, fig.height=5}
highlight_endogenous_all <- merge(DGE_results_4h_filtered, endogenous_groups_data, by = "Gene_ID")

ecdf_4h_g2 <- ggplot(highlight_endogenous_all, aes(
  x = s_cas13d_pos_syn_g2_4h_VS_s_cas13d_4h.logFC, color = group
)) +
  stat_ecdf(geom = "line") +
  coord_cartesian(xlim = c(-0.5, 2)) +
  scale_color_manual(values = highlight_colors) +
  theme_pubr() +
  labs(x = "RNA Log2 FC cth1 KD gRNA2 vs cas13d alone (4 hpf)",
       title = "KD effect by motif class gRNA2, 4 hpf")

ecdf_4h_g2
ggsave(file.path(figures_dir, "ecdf_4h_endogenous_KD_g2.pdf"), ecdf_4h_g2, width = 7, height = 7)
```

## Wilcoxon tests for KD effects

```{r kd-wilcoxon}
kd_wilcox_g2 <- map_dfr(c("2h", "4h"), function(tp) {
  lfc_col <- paste0("s_cas13d_pos_syn_g2_", tp, "_VS_s_cas13d_", tp, ".logFC")
  wilcox_vs_baseline(
    data.frame(highlight_endogenous_all),
    class_col = "group", baseline = "Rest of genes",
    value_col = lfc_col, alternative = "two.sided"
  ) |> mutate(timepoint = tp, grna = "g2")
})

fwrite(kd_wilcox_g2, file.path(results_dir, "wilcoxon_endogenous_g2.csv"))
```

## Endogenous decay analysis (alpha-amanitin)

```{r endogenous-decay}
santi_endogenous <- santi |> merge(endogenous_groups_data, by = "Gene_ID")

pseudocount <- 0.05

santi_decay_alpha <- santi_endogenous |>
  filter(`Treatment_wt-0_hrs-RNAseq_ribo` > 5) |>
  dplyr::select(Gene_ID, contains("aamanitin"), group) |>
  mutate(
    `Alpha_3hpf/2hpf` = log2((`Treatment_aamanitin-3_hrs-RNAseq_polya` + pseudocount) /
                              (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_4hpf/2hpf` = log2((`Treatment_aamanitin-4_hrs-RNAseq_polya` + pseudocount) /
                              (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_5hpf/2hpf` = log2((`Treatment_aamanitin-5_hrs-RNAseq_polya` + pseudocount) /
                              (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_6hpf/2hpf` = log2((`Treatment_aamanitin-6_hrs-RNAseq_polya` + pseudocount) /
                              (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount)),
    `Alpha_7hpf/2hpf` = log2((`Treatment_aamanitin-7_hrs-RNAseq_polya` + pseudocount) /
                              (`Treatment_aamanitin-2_hrs-RNAseq_polya` + pseudocount))
  ) |>
  dplyr::select(-starts_with("Treatment")) |>
  pivot_longer(cols = starts_with("Alpha"),
               names_to = "timepoint", names_pattern = "Alpha_(.*)",
               values_to = "log2_fc")
```

```{r decay-alpha-ecdf, fig.width=10, fig.height=8}
decay_alpha_all <- ggplot(santi_decay_alpha, aes(x = log2_fc, color = group)) +
  stat_ecdf(geom = "step") +
  coord_cartesian(xlim = c(-3, 3)) +
  scale_color_manual(values = highlight_colors) +
  theme_pubr() +
  facet_wrap(~timepoint)

decay_alpha_all
ggsave(file.path(figures_dir, "endogenous_decay_alpha.pdf"), decay_alpha_all,
       width = 12, height = 10)
```

```{r decay-wilcoxon}
endogenous_alpha_pvals <- map_dfr(c("4hpf/2hpf", "5hpf/2hpf", "6hpf/2hpf"), function(tp) {
  wilcox_vs_baseline(
    dplyr::filter(santi_decay_alpha, timepoint == tp),
    class_col = "group", baseline = "Rest of genes", value_col = "log2_fc"
  ) |> mutate(timepoint = tp)
})

fwrite(endogenous_alpha_pvals, file.path(results_dir, "wilcoxon_endogenous_alpha.csv"))
```

## Save intermediate objects

```{r save}
save(
  endogenous_groups, endogenous_groups_data,
  utr3_all, highlight_endogenous_all,
  santi_endogenous, santi_decay_alpha,
  file = file.path(results_dir, "07_endogenous.RData")
)
```
