---
title: "Step 8: Poly(A) Status Scoring and ECDF Analysis"
author: "Daniel Aldas-Bulos"
date: "2026-07-21"
output: html_document
---

## Purpose

Compute poly(A) status scores as log2(polyA TPM / RiboZero TPM) per gene,
then compare fold changes of this score between KD and control conditions
using ECDFs and Wilcoxon/KS tests. Also filters out genes with zero TPM in
both libraries.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source("00_setup.R")
load(file.path(results_dir, "01_groups.RData"))
load(file.path(results_dir, "03_timecourse.RData"))
load(file.path(results_dir, "07_endogenous.RData"))
```

## Compute poly(A) status scores from mean TPMs


```{r compute-scores}

combined_tpm <- fread(file.path(data_dir, "combined_tpm.csv"))

tpm_long <- combined_tpm %>%
  pivot_longer(cols = -c(Gene_ID, Name),
               names_to = "sample_full",
               values_to = "TPM") %>%
  separate(sample_full, into = c("replicate", "library"), sep = "\\.") %>%
  mutate(sample = str_remove(replicate, "_?[123]$"))

tpm_mean <- tpm_long %>%
  group_by(Gene_ID, Name, sample, library) %>%
  summarise(mean_TPM = mean(TPM), .groups = "drop") %>%
  mutate(sample_lib = paste0(sample, ".", library)) %>%
  dplyr::select(Gene_ID, Name, sample_lib, mean_TPM) %>%
  pivot_wider(names_from = sample_lib, values_from = mean_TPM)

pseudocount <- 0.05

# Compute polyA status:
polyA_cols <- grep("\\.polyA$", colnames(tpm_mean), value = TRUE)
samples    <- str_remove(polyA_cols, "\\.polyA$")

score_cols <- map_dfc(samples, function(s) {
  polyA_val <- tpm_mean[[paste0(s, ".polyA")]]
  ribo_val  <- tpm_mean[[paste0(s, ".ribo")]]
  score     <- log2((polyA_val + pseudocount) / (ribo_val + pseudocount))
  tibble(!!paste0("polyA_status.", s) := score)
})

polyA_status <- bind_cols(
  tpm_mean |> dplyr::select(Gene_ID, Name),
  score_cols
)
```


## PolyA status with Cth1 KD inn zygotic genes

```{r}
polyA_status_groups <- inner_join(polyA_status, endogenous_groups_data |> dplyr::select(Gene_ID, group), by = "Gene_ID")

polyA_status_groups_FC <- polyA_status_groups |>
  mutate(polyA_status_FC.alone_vs_g1 =  polyA_status.s_cas13d_pos_syn_g2_4h - polyA_status.s_cas13d_4h,
         polyA_status_FC.alone_vs_g2 =  polyA_status.s_cas13d_pos_syn_g7_4h - polyA_status.s_cas13d_4h,
         polyA_status_FC.alone_vs_g1_with_alpha = polyA_status.s_cas13d_alfa_g2_4h - polyA_status.s_cas13d_alfa_alone_4h,
         polyA_status_FC.alone_vs_g2_with_alpha = polyA_status.s_cas13d_alfa_g7_4h - polyA_status.s_cas13d_alfa_alone_4h,
         ) |>
  dplyr::select(Gene_ID, Name, group, contains("FC"))
```

```{r}
polyA_status_groups |> 
  ggplot(aes(x = polyA_status.s_cas13d_4h, color = group)) +
    stat_ecdf(geom = "step", linewidth = 0.7) +
    coord_cartesian(xlim = c(0, 4)) +
    scale_color_manual(values = highlight_colors) +
    labs(
         x = "polyA status FC",
         y = "Cumulative fraction") +
    theme_pubr(base_size = 11) +
    theme(legend.position = "bottom",
          legend.title = element_blank())
  
```


```{r}
# FC columns to plot
fc_cols <- c("polyA_status_FC.alone_vs_g1",
             "polyA_status_FC.alone_vs_g2",
             "polyA_status_FC.alone_vs_g1_with_alpha",
             "polyA_status_FC.alone_vs_g2_with_alpha")

# Cleaner titles for each plot
fc_labels <- c("alone vs g1", "alone vs g2",
               "alone vs g1 (with alpha)", "alone vs g2 (with alpha)")

groups_of_interest <- c("one site", "3A 8mer", "2A 8mer", "more than one site")

control_vals_list <- map(fc_cols, ~ polyA_status_groups_FC |> filter(group == "Rest of genes") |> pull(.x))
names(control_vals_list) <- fc_cols
```

Discarting genes that has 0 tpms in both polyA and ribo

```{r}
# Identify genes with 0 TPM in both libraries for the alpha samples
unexpressed_alpha <- tpm_mean %>%
  filter(s_cas13d_alfa_alone_4h.polyA == 0 & s_cas13d_alfa_alone_4h.ribo == 0 |
         s_cas13d_alfa_g2_4h.polyA == 0 & s_cas13d_alfa_g2_4h.ribo == 0 |
         s_cas13d_alfa_g7_4h.polyA == 0 & s_cas13d_alfa_g7_4h.ribo == 0) %>%
  pull(Gene_ID)

df_filtered <- polyA_status_groups_FC %>%
  mutate(
    polyA_status_FC.alone_vs_g1_with_alpha = ifelse(
      Gene_ID %in% unexpressed_alpha, NA,
      polyA_status_FC.alone_vs_g1_with_alpha),
    polyA_status_FC.alone_vs_g2_with_alpha = ifelse(
      Gene_ID %in% unexpressed_alpha, NA,
      polyA_status_FC.alone_vs_g2_with_alpha)
  )


```

```{r}
# Build one ECDF plot per FC column with KS p-values
plots <- imap(fc_cols, function(fc_col, i) {
  
  # KS tests: each group vs Rest of genes
  ks_results <- map_dfr(groups_of_interest, function(grp) {
    grp_vals  <- df_filtered %>% filter(group == grp) %>% pull(fc_col) %>% na.omit()
    ctrl_vals <- df_filtered %>% filter(group == "Rest of genes") %>% pull(fc_col) %>% na.omit()
    test <- suppressWarnings(wilcox.test(grp_vals, ctrl_vals))
    tibble(group = grp, D = test$statistic, p = test$p.value, n = length(grp_vals))
  }) %>%
    mutate(
      padj  = p.adjust(p, method = "BH"),
      label = paste0(group, " (n=", n, ")",
                     ", p=", format.pval(padj, digits = 2))
    )
  
  # Plot
  ggplot(df_filtered, aes(x = .data[[fc_col]], color = group)) +
    stat_ecdf(geom = "step", linewidth = 0.7) +
    coord_cartesian(xlim = c(-0.8, 1)) +
    scale_color_manual(values = highlight_colors) +
    annotate("text",
             x = 1,
             y = seq(0.20, 0.05, length.out = nrow(ks_results)),
             label = ks_results$label,
             hjust = 1, size = 2.8, color = "grey30") +
    labs(title = fc_labels[i],
         x = "polyA status FC",
         y = "Cumulative fraction") +
    theme_pubr(base_size = 11) +
    theme(legend.position = "bottom",
          legend.title = element_blank())
})

plots[[1]]
plots[[2]]
plots[[3]]
plots[[4]]

ggsave(file.path(figures_dir, "polyA_FC_alone_vs_g1_no_alpha.pdf"), plots[[1]], width = 7, height = 7)
ggsave(file.path(figures_dir, "polyA_FC_alone_vs_g2_no_alpha.pdf"), plots[[2]], width = 7, height = 7)
ggsave(file.path(figures_dir, "polyA_FC_alone_vs_g1_with_alpha.pdf"), plots[[3]], width = 7, height = 7)
ggsave(file.path(figures_dir, "polyA_FC_alone_vs_g2_with_alpha.pdf"), plots[[4]], width = 7, height = 7)

```



## Session info

```{r session-info}
sessionInfo()
```
